Communicating causal effect heterogeneity

### 2024-06-25 index.qmd
Revision and restructuring

---
title: Communicating causal effect heterogeneity
shorttitle: Communicating heterogeneity
author: 
  - name: Matti Vuorre
    corresponding: true
    orcid: 0000-0001-5052-066X
    email: mjvuorre@uvt.nl
    affiliation: 
      - ref: 1
  - name: Matthew Kay
    affiliation:
      - ref: 2
  - name: Niall Bolger
    affiliation:
      - ref: 3
affiliations:
  - id: 1
    name: Tilburg University
    department: Department of Social Psychology
  - id: 2
    name: Northwestern University
    department: Computer Science and Communication Studies
  - id: 3
    name: Columbia University
    department: Department of Psychology
abstract: "Advances in experimental, data collection, and analysis methods have brought population variability in psychological phenomena to the fore. Yet, current practices for interpreting such heterogeneity underestimate the uncertainty inevitable in any statistical summary. Causal effect heterogeneity is best thought of as distribution of effects with a mean (average person’s effect) and variance (between-person differences). This average or expected heterogeneity distribution can be further summarized as a specific heterogeneity interval [@bolgerCausalProcessesPsychology2019]. However, because the underlying mean and variance parameters will be estimated with uncertainty in empirical studies, these expected distributions and intervals will underestimate the true range of effect values in the population. Using Bayesian hierarchical models, and with the aid of empirical datasets, we provide a walk-through of tools for effective heterogeneity reporting and display. We cover interval, proportion, and ratio measures of heterogeneity and their interpretation. These tools can be a spur to theory building, allowing researchers to widen their focus from population averages to population heterogeneity in psychological phenomena."
keywords: [heterogeneity, uncertainty, variation, hierarchical model, statistics]
bibliography: bibliography.bib
mainfont: Libertinus Serif
numbered-lines: true
floatsintext: true
format:
  apaquarto-pdf: 
    documentmode: doc
    keep-tex: true
    link-citations: true
  apaquarto-docx: default
---

@@

# Packages
library(janitor)
library(scales)
library(ellipse)
library(cmdstanr)
library(readxl)
library(lme4)
library(ggpp)
library(patchwork)
library(latex2exp)
library(ggdist)
library(ggstance)
library(tidybayes)
library(distributional)
library(posterior)
library(parameters)
library(brms)
library(tidyverse)
library(knitr)

# Output options
opts_chunk$set(
  fig.width = 6,
  fig.align = "center",
  eval = TRUE,
  cache = TRUE,
  warning = FALSE,
  error = FALSE,
  message = FALSE
)
options(digits = 2)

@@

When building and testing theories, psychologists have long focused on asking whether an effect exists or not, and what its magnitude might be. Yet, establishing that an independent variable affects a dependent variable, possibly to some specific extent, may not be a sufficient description of the phenomenon if the effect varies appreciably from one treatment unit (e.g. person) to another. The relevance of such variation in the effect, or *heterogeneity*, for theory development is recognized yet typically insufficiently described in the empirical literature [@bolgerCausalProcessesPsychology2019; @brandCausalEffectHeterogeneity2013; @gricePersonsEffectSizes2020; @richters2021].

One reason for the scarcity of reporting and sufficiently interpreting heterogeneity is that psychologists still commonly analyze data with models that obscure its assessment, such as traditional ANOVA [@bolgerCausalProcessesPsychology2019]. However, more informative modeling is not the only challenge: Although more informative hierarchical (or multilevel, mixed-effects) models are becoming widespread, most users do not yet have the conceptual and practical tools to benefit from the greater explanatory power such models afford.

When person-to-person variability is modeled and reported, those descriptions often focus on point estimates [@bolgerCausalProcessesPsychology2019], sample statistics [@beyensEffectSocialMedia2020; @vuorreThreeObjectionsNovel2022; @gricePersonsEffectSizes2020], graphical displays that don't yield numerical estimates of hypothetical data-generating mechanisms[@beck2022], or quantities such as the standard deviation of person-specific parameters [@bartosFairCoinsTend2023]. These, as we will show, provide an incomplete picture of variation: If (e.g.) a treatment effect is found for 60% or participants in a sample but the uncertainty inherent in that percentage is not communicated, we cannot make inferential conclusions about the effect's prevalence in the population. Therefore, to communicate heterogeneity effectively we need not only meaningful measures of it, but also effective methods for describing uncertainty in them. Our goal in this paper is to address this challenge by illustrating measures of heterogeneity and how to communicate it, both numerically and graphically, in ways that take uncertainty into account.

<!-- Fix quarto rendering problem in documentmode: jou. -->

<!-- \clearpage -->

<!-- Fix -->

Our plan is as follows. First, we review established methods for estimating and communicating expected heterogeneity of causal effects in the population using an example dataset from social psychology and describe additional ways in which model parameters can be transformed to describe distributions of causal effects in the population. We review the concepts and computations underlying three heterogeneity metrics: Individual effect’s mean and standard deviation; their heterogeneity interval; and the prevalence proportion. We then move beyond summarizing expected degrees of heterogeneity to describing distributions of plausible degrees of heterogeneity. Finally, we extend these methods to compare heterogeneity across different populations using an example dataset from cognitive psychology.

@@

In this section, we replicate @bolgerCausalProcessesPsychology2019's analysis, using their openly available data. We first wrangled the data as in @bolgerCausalProcessesPsychology2019, which led to a sample of `r number(nrow(dat), big.mark = ",")` trials from `r length(unique(dat$person))` participants that were endorsed as self-descriptive. (We present our code in the online analysis supplement.) We show a sample of these data in @tbl-dat1.

```{r}
#| label: tbl-dat1
#| tbl-cap: First six rows of example dataset 1.
#| apa-note: Reaction time is log-transformed.

dat |>
  head() |>
  rename(`Reaction time` = rt) |> 
  rename_with(~str_to_sentence(.x)) |> 
  kable()
```

Then, we estimated the same statistical model (Model 1; equations 1.1-1.3). We modeled the log-transformed reaction time (logRT) of person $j$ on trial $i$ as a random draw from a normal distribution with mean $\eta$ (*eta*), which could differ between trials $i$ and individuals $j$, and standard deviation $\sigma$ (*sigma*), which we assumed constant across individuals and trials as indicated by the lack of subscripts:

```{=tex}
\begin{equation}
\tag{1.1}
\text{logRT}_{ij} \sim \operatorname{Normal}\left(\eta_{ij}, \sigma^2\right).
\end{equation}
```
(We recognize that there are better alternatives to modeling the log-transformed RTs as normal, but those are outside the scope of this manuscript.) Then, we specified a model of the mean of the logRT distribution ($\eta_{ij}$) such that the regression coefficients captured our substantive questions:

```{=tex}
\begin{equation}
\tag{1.2}
\eta_{ij} = \beta_0 + \gamma_{0j} + \left(\beta_1 + \gamma_{1j}\right)\text{V}_{ij}.
\end{equation}
```
This equation includes two sets of parameters: The first set contains $\beta_0$ (*beta*), the intercept, and $\beta_1$, the slope or effect of valence (V). Parameters in this set do not have subscripts: In the frequentist tradition, they are considered constants---not modelled on covariates---and typically referred to as "fixed" parameters [e.g. @raudenbushHierarchicalLinearModels2002]. The second set of parameters, $\gamma_{0j}$ (*gamma*) and $\gamma_{1j}$, have the subscript $j$ to indicate that they are person-specific deviations from the average intercept and slope, respectively. That is, $\beta_0 + \gamma_{01}$ is the intercept (average reaction time) for person $j=1$. In frequentist texts, these are typically called "random" parameters because they are modeled as varying randomly according to a specified distribution. Following standard multilevel modeling assumptions, we model $\gamma_0$ and $\gamma_1$ as multivariate normal distributed:

```{=tex}
\begin{equation}
\tag{1.3}
\begin{bmatrix} 
  \gamma_0 \\ \gamma_1
\end{bmatrix} \sim 
\operatorname{MVN}\left(
  \begin{bmatrix} 0 \\ 0 \end{bmatrix}, 
  \begin{pmatrix} 
    \tau_0 & \\ 
    \rho &\tau_1 
  \end{pmatrix}
\right).
\end{equation}
```
In this equation, we assume that the person-specific deviations $\gamma_0$ and $\gamma_1$ have means of zero (because the means are added to them in equation 1.2), standard deviations $\tau$ (*tau*), and a correlation $\rho$ (*rho*). Perhaps confusingly, $\tau$s and $\rho$ are also sometimes called random effects because they describe random (co)variations of the person-specific effects. To be clear, despite this naming convention they are features of the population, not of any one group or individual.

@@

We show a conventional summary of this model's estimated parameters in @tbl-lmer. For the average person, the estimated effect of positive valence on logRT is `r mu1[1]` log seconds, with a 95% confidence interval (CI) extending from `r mu1[2]` to `r mu1[3]`. The estimated standard deviation of valence effects in the population is `r sd1[1]` log seconds. The lme4 software package does not report a standard error or CI for (co)variance parameters by default, and we therefore calculated it by bootstrapping. The resulting 95% bootstrap CI of the valence effect's standard deviation was \[`r sd1[2]`, `r sd1[3]`\].

@@

```{r}
#| label: fig-1
#| fig.height: 1.8
#| apa-twocolumn: true
#| fig-cap: Heterogeneity distribution of valence effects and various descriptions of their expected heterogeneity as estimated with Model 1. A. The normal density curve defined by the point estimates of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). Shaded areas represent areas under the normal curve within 1 (dark) and 2 (light) standard deviations of the mean. B. The 90\% Heterogeneity Interval as represented by a line segment with arrows, and the dark shaded area. C. Proportion of negative valence effects in the population (dark). D. Proportion of valence effects in the population that are within the region of practical equivalence to zero (ROPE; dark).

@@

Rows 2 and 4 in @tbl-lmer define the expected normal distribution of valence effects in the population, visualized in @fig-1 A. In other words, our point estimate of the distribution of valence effects is Normal(`r mu1[1]`, `r sd1[1]`^2^). However, this distribution is an incomplete description of heterogeneity for two reasons. First, it does not incorporate uncertainty in the two determinants of heterogeneity, that is $\beta_1$ and $tau_1$: If they are precisely estimated, i.e. when uncertainty regarding them is negligible, the distribution and any quantities calculated from it characterize the population well. On the other hand, if they are estimated with considerable uncertainty, the distribution or its transformations would not characterize the population well. We return to this key issue below. Second, the distribution or its parameters do not, for many purposes, communicate heterogeneity in clear and actionable terms. Below, we introduce several metrics that e.g. directly describe where a given proportion of the slopes are expected to fall.

@@

To calculate a heterogeneity interval, we first specify the desired probability limits. For a 90% interval, we use .05 and .95, which together define the central 90% of the distribution. Then, we pass those limits and the estimated mean and standard deviation to the normal quantile function $\Phi^{-1}$ (*phi*, `qnorm()` in R), to get the interval: $HI_{90} = \Phi^{-1}([.05, .95], \beta_1, \tau_1) = \Phi^{-1}([.05, .95],$ `r mu1[1]`, `r sd1[1]`) = \[`r hi90`\]. In words, this function calculates the 0.05 and 0.95 quantiles of the normal distribution defined by the mean's ($\beta_1$) and standard deviation's ($\tau_1$) point estimates: We expect 90% of valence effects in the population to fall within this interval. We illustrate this interval in @fig-1 B.

@@

The above HI summarizes where a given proportion of individuals' effects lie. In contrast, some applications might find it more informative to summarize proportions of effects above or below some critical value, or within some critical range. For example, we might ask "What proportion of individuals in the population respond faster to positively valenced words?" In other words, we ask a question of *prevalence*: What proportion of the heterogeneity distribution is below zero? We label this quantity $p^-$ for proportion of population with negative effects.

@@

However, using zero as a critical value might not be sufficiently informative, especially when theory allows specifying a smallest effect size of interest, or what is known as a region of practical equivalence [ROPE, @anvariUsingAnchorbasedMethods2021; @lakensEquivalenceTestingPsychological2018; @kruschkeBayesianDataAnalysis2017; @kruschkeDoingBayesianData2014]. In common applications, ROPE is used to statistically infer whether an estimated parameter, such as the effect of valence on logRT for the average person, is practically significant. But we can equally well use a theory-informed region of parameter values to describe and make inferences about the heterogeneity distribution of this effect in the population.

For example, let us imagine that a theory states that valence effects in the interval \[-0.1, 0.1\] are practically equivalent to zero. To calculate, we can again use the normal cumulative distribution function to calculate the proportion of individuals in the population whose valence effect falls within this interval or region of practical equivalence: $p^{ROPE} = \Phi(0.1, \beta_1, \tau_1) - \Phi(-0.1, \beta_1, \tau_1)$ = `r p_rope`. In words, `r p_rope` of the population is expected to have valence effects that are practically equivalent to zero. Note that this statement's validity critically depends on the chosen interval. We visualize this probability in @fig-1 D.

@@

[@bolgerCausalProcessesPsychology2019, p.609] suggest as a rule of thumb that heterogeneity can be deemed noteworthy when the ratio of the standard deviation to the average effect is 0.25 or greater: A ratio greater than 1/4 indicates a $HI_{95}$ whose limits extend to effects one-half and one-and-a-half times that of the average effect. With these data and model, the ratio $\frac{\tau_1}{\beta_1}$ is `r ratio`, suggesting that the degree of heterogeneity in valence effects is noteworthy. While this heuristic can sometimes be useful, we urge users to apply domain-specific knowledge when considering critical values or thresholds whenever possible.

@@

Assessments of heterogeneity involve combining information about fixed and random effects; to fully incorporate inferential uncertainty then requires accounting for their uncertainties in combination. Probabilistic, that is, Bayesian methods are uniquely able to address this challenge. Modern Bayesian methods, by obtaining draws from the joint posterior distribution of all model parameters presumed to underlie the observed data, allow incorporating posterior uncertainty in combinations of parameters such as the ones highlighted above [@gelmanBayesianDataAnalysis2013; @kruschkeBayesianDataAnalysis2017; @kruschkeDoingBayesianData2014; @mcelreathStatisticalRethinkingBayesian2020]. For typical scenarios, Bayesian models are as easy to use as their maximum likelihood counterparts [@burknerBrmsPackageBayesian2017; @burknerAdvancedBayesianMultilevel2018].

@@

The output of Bayesian computations is the multivariate posterior probability distribution of the model's parameters. However, closed-form solutions are not available for multivariate posterior distributions of many important types of statistical models. Therefore in practice modern Bayesian methods rely on algorithms that yield many random draws from the multivariate posterior distribution [@gelmanBayesianDataAnalysis2013; @ravenzwaaijSimpleIntroductionMarkov2016]. These draws can then be used to calculate, summarize, and visualize any desired quantity of the multivariate posterior such as means, variances, correlations, proportions above or below zero, and so on. @tbl-samples-1 illustrates this, showing six random draws of the posteriors of $\beta_1$ and $\tau_1$ (rows). We then computed their ratio in the third column, which then represents (draws from) the ratio's posterior distribution and can be summarized, visualized, etc.

@@

In practice, one obtains (for example) 4,000 samples from the posterior distribution using Markov Chain Monte Carlo algorithms [e.g. @standevelopmentteamStanModelingLanguage2023] through accessible software [e.g., @burknerBrmsPackageBayesian2017], and then summarizes them using familiar data processing techniques [e.g. @dplyr2023; @rcoreteamLanguageEnvironmentStatistical2023]. Here, we used the R package brms [@burknerBrmsPackageBayesian2017; @burknerAdvancedBayesianMultilevel2018] to specify the model and then draw random samples from its posterior distribution. The MCMC estimation algorithm completed in about 5 seconds on a modern laptop. We then assessed the estimation algorithm convergence graphically and numerically, and model adequacy using a graphical posterior predictive check [@gelmanBayesianDataAnalysis2013].

@@

Armed with the Bayesian draws (@tbl-fit-1), we can now return to the question of the distribution of valence effects in the population. We now have 4,000 samples from this heterogeneity distribution's posterior distribution. We first redraw the expected heterogeneity distribution from @fig-1 using the posterior mean values of $\beta_1$ and $\tau_1$ in @fig-2 A (thick dark blue curve). Superimposed on that normal density curve are heterogeneity distributions calculated from 100 random posterior draws of $\beta_1$ and $\tau_1$. From these curves we can see that the true distribution of valence effects might well be less or more heterogeneous than is suggested by the expectation (point estimates).

@@

```{r}
#| label: fig-2
#| fig-width: 4
#| fig-height: 1.8
#| fig-cap: Bayesian estimates of the heterogeneity distribution of valence effects.
#| apa-note: "A: Probability density function (PDF) curves. The thick line is the same expected PDF of valence effects from Figure 1. Thin lines show 100 PDFs calculated from random draws of $\\beta_1$ and $\\tau_1$ that collectively illustrate the uncertainty in the distribution's location and spread. B: Cumulative density function (CDF) curves, annotated as A. C. Histogram of (draws from) the posterior distribution of $p^-$, which correspond to CDF segments below zero: This value describes the proportion of individuals with negative valence effects on logRT. Point and interval represents the posterior mean and 95\\%CI. Solid and dotted vertical lines in A and B highlight x-axis values of 0, and -0.1 and 0.1, respectively."

@@

@fig-intervals A further suggests that communicating the two uncertainty intervals of a heterogeneity interval is not only cumbersome, but also ignores potential correlations between the heterogeneity interval endpoints' posterior distributions. For these reasons, although the HI can be a useful summary, we occasionally favor [e.g., @vuorreAffectiveUpliftVideo2024] the scalar descriptors discussed below.

>>>>

---
title: Communicating causal effect heterogeneity
running-head: Communicating heterogeneity
author: 
  - name: Matti Vuorre
    orcid: 0000-0001-5052-066X
    email: mjvuorre@uvt.nl
    corresponding: true
    affiliation: 
      - ref: 1
  - name: Matthew Kay
    orcid: 0000-0001-9446-0419
    affiliation:
      - ref: 2
  - name: Niall Bolger
    orcid: 0000-0001-8698-8117
    affiliation:
      - ref: 3
affiliations:
  - id: 1
    name: Tilburg University
    department: Department of Social Psychology
  - id: 2
    name: Northwestern University
    department: Computer Science and Communication Studies
  - id: 3
    name: Columbia University
    department: Department of Psychology
abstract: |
  Advances in experimental, data collection, and analysis methods have brought population variability in psychological phenomena to the fore. Yet, current practices for interpreting such heterogeneity do not appropriately treat the uncertainty inevitable in any statistical summary. Heterogeneity is best thought of as a distribution of features with a mean (average person’s effect) and variance (between-person differences). This expected heterogeneity distribution can be further summarized e.g. as a heterogeneity interval [@bolgerCausalProcessesPsychology2019]. However, because empirical studies estimate the underlying mean and variance parameters with uncertainty, the expected distribution and interval will underestimate the actual range of plausible effects in the population. Using Bayesian hierarchical models, and with the aid of empirical datasets from social and cognitive psychology, we provide a walk-through of effective heterogeneity reporting and display tools that appropriately convey measures of uncertainty. We cover interval, proportion, and ratio measures of heterogeneity and their estimation and interpretation. These tools can be a spur to theory building, allowing researchers to widen their focus from population averages to population heterogeneity in psychological phenomena.
keywords: [heterogeneity, uncertainty, variation, hierarchical model, statistics]
authornote: |
  This manuscript is not yet peer-reviewed.
bibliography: bibliography.bib
citeproc: true
csl: apa.csl
---

@@

# Packages
library(janitor)
library(scales)
library(ellipse)
library(tinytable)
library(cmdstanr)
library(readxl)
library(lme4)
library(ggpp)
library(patchwork)
library(latex2exp)
library(ggdist)
library(ggstance)
library(tidybayes)
library(distributional)
library(posterior)
library(parameters)
library(brms)
library(tidyverse)
library(knitr)

# Output options
opts_chunk$set(
  eval = TRUE,
  cache = TRUE,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  echo = FALSE
)
options(digits = 2)
options(tinytable_tt_digits = 3)

@@

When building and testing theories, psychologists have long focused on asking whether an effect exists and what its magnitude might be. Yet, establishing that an independent variable affects a dependent variable, possibly to some specific extent, may not be a sufficient description of the phenomenon if the effect varies appreciably from one treatment unit (e.g. person) to another. The relevance of such variation in the effect, or *heterogeneity*, for theory development is recognized yet typically insufficiently described in the empirical literature [@bolgerCausalProcessesPsychology2019; @brandCausalEffectHeterogeneity2013; @gricePersonsEffectSizes2020; @richters2021].

One reason for the scarcity of reporting and sufficiently interpreting heterogeneity is that psychologists still commonly analyze data with models that obscure its assessment, such as traditional ANOVA [@bolgerCausalProcessesPsychology2019]. However, more informative modeling is not the only challenge: Although more informative hierarchical (or multilevel, mixed-effects [@gelmanDataAnalysisUsing2007]) models are becoming widespread, many users do not yet have the conceptual and practical tools to benefit from the greater explanatory power such models afford.

When person-to-person variability is modeled and reported, those descriptions often focus on point estimates [@bolgerCausalProcessesPsychology2019], sample statistics [@beyensEffectSocialMedia2020; @vuorreThreeObjectionsNovel2022; @gricePersonsEffectSizes2020], graphical displays that don't yield numerical estimates of hypothetical data-generating mechanisms [@beck2022], or quantities such as the standard deviation of person-specific parameters [@bartosFairCoinsTend2023]. These, as we will show, provide an incomplete picture of variation that is sometimes difficult to interpret: If (e.g.) a treatment effect is found for 60% or participants in a sample but the uncertainty inherent in that percentage is not communicated, we cannot make inferential conclusions about the effect's prevalence in the population. Therefore, to communicate heterogeneity effectively we need not only meaningful measures of it, but also effective methods for describing the associated uncertainties. Our goal in this paper is to address this challenge by illustrating measures of heterogeneity and how to communicate them, both numerically and graphically, in ways that take uncertainty into account.

Our plan is as follows. First, we review established methods for estimating and communicating expected heterogeneity of causal effects in the population using an example dataset from social psychology. We then describe additional ways in which model parameters can be transformed to describe distributions of causal effects. We review the concepts and computations underlying three heterogeneity metrics: The effect's mean and standard deviation in the population; the heterogeneity interval; and the prevalence proportion. Second, we move beyond summarizing expected degrees of heterogeneity that lack information about uncertainty to describing distributions of plausible degrees of heterogeneity. Such uncertainty distributions of population feature distributions are natural components of Bayesian hierarchical models and afford efficient tools for describing distributional uncertainty. Finally, we extend these methods to compare heterogeneity across different populations using an example dataset from cognitive psychology.

@@

In this section, we replicate @bolgerCausalProcessesPsychology2019's analysis, using their openly available data. We first wrangled the data as in @bolgerCausalProcessesPsychology2019, which led to a sample of `r number(nrow(dat), big.mark = ",")` trials from `r length(unique(dat$person))` participants that were endorsed as self-descriptive. Our online analysis supplement (<https://osf.io/yp2gq>) includes the complete code to reproduce this manuscript and computations therein. We show a sample of these data in @tbl-dat1.

```{r}
#| label: tbl-dat1
#| tbl-cap: First six rows of example dataset 1 [@bolgerCausalProcessesPsychology2019].

dat |>
  head() |>
  rename(`log(Reaction time)` = rt) |> 
  rename_with(~str_to_sentence(.x)) |> 
  kable()
```

Then, we estimated the same statistical model (@eq-m1-1 - [-@eq-m1-3]). We modeled the log-transformed reaction time of person $j$ on trial $i$ as a random draw from a normal distribution with mean $\eta$ (*eta*), which could differ between trials $i$ and individuals $j$, and standard deviation $\sigma$ (*sigma*), which we assumed constant across individuals and trials as indicated by the lack of subscripts:

$$
\text{logRT}_{ij} \sim \operatorname{Normal}\left(\eta_{ij}, \sigma^2\right).
$$ {#eq-m1-1}

(We recognize that there are better alternatives to modeling the log-transformed RTs as normal, but those are outside the scope of this manuscript.) Then, we specified a model of the mean of the logRT distribution ($\eta_{ij}$) such that the regression coefficients captured our substantive questions:

$$
\eta_{ij} = \beta_0 + \gamma_{0j} + \left(\beta_1 + \gamma_{1j}\right)\text{V}_{ij}.
$$ {#eq-m1-2}

This equation includes two sets of parameters: The first set contains $\beta_0$ (*beta*), the intercept, and $\beta_1$, the slope or effect of valence (V). Parameters in this set do not have subscripts: In the frequentist tradition, they are considered constants---not modelled on covariates---and typically referred to as "fixed" parameters [e.g. @raudenbushHierarchicalLinearModels2002]. The second set of parameters, $\gamma_{0j}$ (*gamma*) and $\gamma_{1j}$, have the subscript $j$ to indicate that they are person-specific deviations from the average intercept and slope, respectively. That is, $\beta_0 + \gamma_{01}$ is the intercept (average reaction time) for person $j=1$. In frequentist texts, these are typically called "random" parameters because they are modeled as varying randomly according to a specified distribution. Following standard multilevel modeling assumptions, we model $\gamma_0$ and $\gamma_1$ as multivariate normal distributed:

$$
\begin{bmatrix} 
  \gamma_0 \\ \gamma_1
\end{bmatrix} \sim 
\operatorname{MVN}\left(
  \begin{bmatrix} 0 \\ 0 \end{bmatrix}, 
  \begin{pmatrix} 
    \tau_0 & \\ 
    \rho &\tau_1 
  \end{pmatrix}
\right).
$$ {#eq-m1-3}

In this equation, we assume that the person-specific deviations $\gamma_0$ and $\gamma_1$ have means of zero (because the means are added to them in equation 1.2), standard deviations $\tau$ (*tau*), and a correlation $\rho$ (*rho*). Perhaps confusingly, $\tau$s and $\rho$ are also sometimes called random effects because they describe random (co)variations of the person-specific effects. To be clear, despite this naming convention they are features of the population, not of any one group or individual.

@@

We show a conventional summary of this model's estimated parameters in @tbl-lmer. For the average person, the estimated effect of positive valence on logRT is `r mu1[1]` log seconds, with a 95% confidence interval (CI) extending from `r mu1[2]` to `r mu1[3]`. The estimated standard deviation of valence effects in the population is `r sd1[1]` log seconds. The lme4 software package does not report a standard error or CI for (co)variance parameters by default, and we therefore calculated it by bootstrapping. The resulting 95% bootstrap CI of the valence effect's standard deviation was [`r sd1[2]`, `r sd1[3]`].

@@

```{r}
#| label: fig-1
#| fig-height: 2
#| fig-width: 7
#| fig-cap: Heterogeneity distribution of valence effects and various descriptions of their expected heterogeneity as estimated with Model 1. A. The normal density curve defined by the point estimates of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). Shaded areas represent areas under the normal curve within 1 (dark) and 2 (light) standard deviations of the mean. B. The 90\% Heterogeneity Interval as represented by a line segment with arrows, and the dark shaded area. C. Proportion of negative valence effects in the population (dark). D. Proportion of valence effects in the population that are within the region of practical equivalence to zero (ROPE; dark).

@@

Rows 2 and 4 in @tbl-lmer define the expected normal distribution of valence effects in the population, visualized in @fig-1 A. In other words, our point estimate of the distribution of valence effects is Normal(`r mu1[1]`, `r sd1[1]`^2^). However, this distribution is an incomplete description of heterogeneity for two reasons. First, it does not incorporate uncertainty in the two determinants of heterogeneity, that is $\beta_1$ and $\tau_1$: If they are precisely estimated, i.e. when uncertainty regarding them is negligible, the distribution and any quantities calculated from it characterize the population well. On the other hand, if they are estimated with considerable uncertainty, the distribution or its transformations would not characterize the population well. We return to this key issue below. Second, the distribution or its parameters do not, for many purposes, communicate heterogeneity in clear and actionable terms. Below, we introduce several metrics that directly describe e.g. where a given proportion of the slopes are expected to fall.

@@

To calculate a heterogeneity interval, we first specify the desired probability limits. For a 90% interval, we use .05 and .95, which together define the central 90% of the distribution. Then, we pass those limits and the estimated mean and standard deviation to the normal quantile function $\Phi^{-1}$ (*phi*, `qnorm()` in R), to get the interval: $HI_{90} = \Phi^{-1}([.05, .95], \beta_1, \tau_1) = \Phi^{-1}([.05, .95],$ `r mu1[1]`, `r sd1[1]`) = [`r hi90`]. In words, this function calculates the 0.05 and 0.95 quantiles of the normal distribution defined by the mean's ($\beta_1$) and standard deviation's ($\tau_1$) point estimates: We expect 90% of valence effects in the population to fall in the [`r hi90`] interval. We illustrate this interval in @fig-1 B.

@@

The above HI summarizes where a given proportion of individuals' effects in the population are. In contrast, some applications might find it more informative to summarize proportions of effects above or below some critical value, or within some critical range. For example, we might ask "What proportion of individuals in the population respond faster to positively valenced words?" In other words, we ask a question of *prevalence*: What proportion of the heterogeneity distribution is below zero? We label this quantity $p^-$ for proportion of population with negative effects.

@@

However, using zero as a critical value might not be sufficiently informative, especially when theory allows specifying a smallest effect size of interest, or what is known as a region of practical equivalence [ROPE, @anvariUsingAnchorbasedMethods2021; @lakensEquivalenceTestingPsychological2018; @kruschkeBayesianDataAnalysis2017; @kruschkeDoingBayesianData2014]. In common applications, ROPE is used to statistically infer whether an estimated parameter, such as the effect of valence on logRT for the average person, is practically significant. But we can equally well use a theory-informed region of effect sizes to describe and make inferences about the heterogeneity distribution of this effect in the population.

For example, let us imagine that a theory states that valence effects in the interval [-0.1, 0.1] are practically equivalent to zero. To calculate, we can again use the normal cumulative distribution function to calculate the proportion of individuals in the population whose valence effect falls within this interval or region of practical equivalence: $p^{ROPE} = \Phi(0.1, \beta_1, \tau_1) - \Phi(-0.1, \beta_1, \tau_1)$ = `r p_rope`. In words, `r p_rope` of the population is expected to have valence effects that are practically equivalent to zero. Note that this statement's validity critically depends on the chosen interval's theoretical validity. We visualize this probability in @fig-1 D.

@@

[@bolgerCausalProcessesPsychology2019, p.609] suggest as a rule of thumb that heterogeneity can be deemed noteworthy when the ratio of the standard deviation to the average effect is 0.25 or greater: A ratio greater than 1/4 implies a $HI_{95}$ whose limits extend to effects one-half and one-and-a-half times that of the average effect. With these data and model, the ratio $\frac{\tau_1}{\beta_1}$ is `r ratio`, suggesting that the degree of heterogeneity in valence effects is noteworthy. While this heuristic can sometimes be useful, we urge users to apply domain-specific knowledge when considering critical values or thresholds whenever possible.

@@

Assessments of heterogeneity involve combining information about fixed and random effects; to fully incorporate inferential uncertainty then requires accounting for their joint uncertainties. Probabilistic, that is, Bayesian methods are uniquely able to address this challenge. Modern Bayesian methods, by obtaining draws from the joint posterior distribution of all model parameters presumed to underlie the observed data, allow incorporating posterior uncertainty in combinations of parameters such as the ones highlighted above [@gelmanBayesianDataAnalysis2013; @kruschkeBayesianDataAnalysis2017; @kruschkeDoingBayesianData2014; @mcelreathStatisticalRethinkingBayesian2020]. For typical scenarios, Bayesian models are as easy to use as their maximum likelihood counterparts [@burknerBrmsPackageBayesian2017; @burknerAdvancedBayesianMultilevel2018].

@@

The output of Bayesian computations is the multivariate posterior probability distribution of the model's parameters. However, closed-form solutions are not available for multivariate posterior distributions of many important types of statistical models. Therefore, in practice modern Bayesian methods rely on algorithms that yield many random draws from the multivariate posterior distribution [@gelmanBayesianDataAnalysis2013; @ravenzwaaijSimpleIntroductionMarkov2016]. These draws can then be used to calculate, summarize, and visualize any desired quantity of the multivariate posterior such as means, variances, correlations, proportions above or below zero, and so on. @tbl-samples-1 illustrates this, showing six random draws of the posteriors of $\beta_1$ and $\tau_1$ (rows). We then computed their ratio in the third column, which then represents (draws from) the ratio's posterior distribution and can be summarized, visualized, etc.

@@

In practice, one obtains (for example) 4,000 samples from the posterior distribution using Markov Chain Monte Carlo algorithms [e.g. @standevelopmentteamStanModelingLanguage2023] through accessible software [e.g., @burknerBrmsPackageBayesian2017], and then summarizes them using familiar data processing techniques [e.g. @dplyr2023; @rcoreteamLanguageEnvironmentStatistical2023]. Here, we used the R package brms [@burknerBrmsPackageBayesian2017; @burknerAdvancedBayesianMultilevel2018] to specify the model and then draw random samples from its posterior distribution. The MCMC estimation algorithm completed in about 5 seconds on a modern laptop. We then assessed the estimation algorithm convergence graphically and numerically, and model adequacy using a graphical posterior predictive check [@gelmanBayesianDataAnalysis2013]. (These, and other details, are presented in our supplementary online analyses.)

@@

Armed with the Bayesian draws (@tbl-fit-1), we can now return to the question of the distribution of valence effects in the population. We now have 4,000 samples from this heterogeneity distribution's posterior distribution. Effectively, then, we have among other quantities 4,000 samples from the heterogeneity distribution's posterior distribution. We first redraw the expected heterogeneity distribution from @fig-1 using the posterior mean values of $\beta_1$ and $\tau_1$ in @fig-2 A (thick dark blue curve). Superimposed on that normal density curve are heterogeneity distributions calculated from 100 random posterior draws of $\beta_1$ and $\tau_1$. From these curves we can see that the true distribution of valence effects might well be less or more heterogeneous than is suggested by the expectation (point estimates).

@@

```{r}
#| label: fig-2
#| fig-width: 5.4
#| fig-height: 2.2
#| fig-cap: "Bayesian estimates of the heterogeneity distribution of valence effects. A: Probability density function (PDF) curves. The thick line is the same expected PDF of valence effects from Figure 1. Thin lines show 100 PDFs calculated from random draws of $\\beta_1$ and $\\tau_1$ that collectively illustrate the uncertainty in the distribution's location and spread. B: Cumulative density function (CDF) curves, annotated as A. C. Histogram of (draws from) the posterior distribution of $p^-$, which correspond to CDF segments below zero: This value describes the proportion of individuals with negative valence effects on logRT. Point and interval represents the posterior mean and 95\\%CI. Solid and dotted vertical lines in A and B highlight x-axis values of 0, and -0.1 and 0.1, respectively."

@@

```{r}
#| label: fig-intervals
#| fig-height: 2.4
#| fig-cap: Bayesian estimates of the 90% heterogeneity interval of valence effects. A. Scatterplot of 4,000 posterior samples of the lower (x-axis) and upper (y-axis) limits of the $HI_{90}$. B. Histograms of 4,000 samples of the HI90 lower (left) and upper (right) limits with their posterior means and 95%CIs as points and intervals. C. 100 random samples from the $HI_{90}$ posterior distribution, with the posterior mean heterogeneity interval superimposed in a darker shade of blue.

p2 <- post1 |>
  mutate(
    hi_low = b_valence1 + qnorm(0.05) * sd_person__valence1,
    hi_high = b_valence1 + qnorm(0.95) * sd_person__valence1
  ) |>
  select(hi_low, hi_high) |>
  pivot_longer(everything()) |>
  ggplot(aes(value, group = name)) +
  scale_y_continuous(
    expand = expansion(c(0, 0.1))
  ) +
  scale_x_continuous(
    "Valence effect"
  ) +
  stat_histinterval(
    breaks = 100,
    fill = alpha("dodgerblue1", .33),
    color = "dodgerblue4",
    .width = c(.95),
    justification = .01
  ) +
theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
